---
title: Tidying STM with tidytext
author: Ian Adams
date: '2020-06-23'
categories:
  - R
  - research
tags:
  - police
  - R Markdown
  - STM
  - survey
slug: tidying-stm-with-tidytext
lastmod: '2020-06-23T12:07:06-06:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

## Libraries
```{r message=FALSE, warning=FALSE}
library(tidytext)
library(ggthemes)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(scales)
```

## Load Previous STM Objects
I have previously run `stm` models for topics ranging from 3 to 25. Based on the fit indices, a six-topic model was selected. I am not showing that analysis here, but instead loading the saved `stm` objects (the model and the 'out' object).
```{r remedy001, include=FALSE}

# Load objects if you want to skip the first half of specifying the model
## Theta df
load("D:/Google Drive (SG)/R/Dissertation/Dissertation Analysis/STM/Stored Results/DisswithTheta_bwc_only.Rdata")
## Ksearch result object
load("D:/Google Drive (SG)/R/Dissertation/Dissertation Analysis/STM/Stored Results/kResult3_25_bwc_only.Rdata")
## STM object
load("D:/Google Drive (SG)/R/Dissertation/Dissertation Analysis/STM/Stored Results/modelfit6_bwc_only.Rdata")
## Out object
load("D:/Google Drive (SG)/R/Dissertation/Dissertation Analysis/STM/Stored Results/out_bwc_only.Rdata")


```

## Tidy the stm model

I am using the `tidytext` package from Julia Silge, which she [demonstrates in a blog post here](https://juliasilge.com/blog/evaluating-stm/). 

```{r remedy002}

td_beta <- tidy(modelfit6)
td_beta

td_gamma <- tidy(modelfit6, matrix = "gamma",
                 document_names = rownames(out$meta))

td_gamma

```

## Plot Topic Relevance

```{r remedy003}

top_terms <- td_beta %>%
  arrange(beta) %>%
  group_by(topic) %>%
  top_n(7, beta) %>%
  arrange(-beta) %>%
  select(topic, term) %>%
  summarise(terms = list(term)) %>%
  mutate(terms = map(terms, paste, collapse = ", ")) %>% 
  unnest()

gamma_terms <- td_gamma %>%
  group_by(topic) %>%
  summarise(gamma = mean(gamma)) %>%
  arrange(desc(gamma)) %>%
  left_join(top_terms, by = "topic") %>%
  mutate(topic = paste0("Topic ", topic),
         topic = reorder(topic, gamma))

gamma_terms %>%
  top_n(6, gamma) %>%
  ggplot(aes(topic, gamma, label = terms, fill = topic)) +
  geom_col(show.legend = FALSE) +
  geom_text(hjust = 0.85, nudge_y = 0.0005, size = 3) +
  coord_flip() +
  theme_hc() +
  theme(plot.title = element_text(size = 14)) +
  labs(x = NULL, y = expression(gamma),
       title = "Top Six Topics by Prevalence in the Officer Responses",
       subtitle = "With the top words that contribute to each topic")

```

## Table of Topic Proportions with Top Terms

```{r remedy004}

require(knitr)
gamma_terms %>%
  select(topic, gamma, terms) %>%
  kable(digits = 3, 
        col.names = c("Topic", "Expected topic proportion", "Top 6 terms"))

```
