---
title: Counting "Select All That Apply" Questions in Qualtrics
author: ian-adams
date: '2021-08-25'
slug: counting-select-all-that-apply-questions-in-qualtrics-using-R
categories:
  - R
  - stats
  - research
tags:
  - Demo
  - R Markdown
  - qualtrics
  - stats
  - survey
  - research
subtitle: ''
summary: ''
authors: []
lastmod: '2021-08-25T09:17:35-06:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r library, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(tidyverse)
library(readxl)
library(tidyr)
library(here)
library(janitor)

```

```{r data-munge, include=FALSE}

dat_raw <- read_excel(here("assets", "Student_Thesis.xlsx")) %>%
  clean_names()

dat <- dat_raw %>% select(
  grammarly_awareness_1:grammarly_awarness_4
)

## Recoding dat_raw$grammarly_awareness_1 into dat_raw$color_1
dat$color_1 <- fct_recode(dat$grammarly_awareness_1,
  "Blue" = "Installed and use occasionally",
  "Green" = "Not tried it",
  "Yellow" = "Other levels of awareness/experience:",
  "Sparkle" = "Use for most of my work",
  "Apple Core Brown" = "Used the anti-plagiarism check feature",
  "Honey Bee" = "WILL NOT USE on my thesis",
  "Moldy Book" = "WILL USE on my thesis"
)

dat$color_2 <- fct_recode(dat$grammary_awareness_2,
  "Blue" = "Installed and use occasionally",
  "Green" = "Not tried it",
  "Yellow" = "Other levels of awareness/experience:",
  "Sparkle" = "Use for most of my work",
  "Apple Core Brown" = "Used the anti-plagiarism check feature",
  "Honey Bee" = "WILL NOT USE on my thesis",
  "Moldy Book" = "WILL USE on my thesis"
)

dat$color_3 <- fct_recode(dat$grammarly_awareness_3,
  "Blue" = "Installed and use occasionally",
  "Green" = "Not tried it",
  "Yellow" = "Other levels of awareness/experience:",
  "Sparkle" = "Use for most of my work",
  "Apple Core Brown" = "Used the anti-plagiarism check feature",
  "Honey Bee" = "WILL NOT USE on my thesis",
  "Moldy Book" = "WILL USE on my thesis"
)

dat$color_4 <- fct_recode(dat$grammarly_awarness_4,
  "Blue" = "Installed and use occasionally",
  "Green" = "Not tried it",
  "Yellow" = "Other levels of awareness/experience:",
  "Sparkle" = "Use for most of my work",
  "Apple Core Brown" = "Used the anti-plagiarism check feature",
  "Honey Bee" = "WILL NOT USE on my thesis",
  "Moldy Book" = "WILL USE on my thesis"
)

dat <- dat %>% select(
  color_1:color_4
)

```


# Qualtrics Messy Data

My friend [Devon Cantwell](https://www.devoncantwell.com/about-devon) reached out with an interesting messy data caused by how Qualtrics produces "select all that apply" variables. For example, in her (mock) survey, she asks students to select all the colors that they personally find attractive from a list. When downloaded from Qualtrics, we get a dataframe that looks like this:

```{r remedy001}

glimpse(dat)

```
So all students pick at least one color, some pick two, but relatively few pick three or four. One thing we might want to know is the first color selected by respondent? That's relatively easy:

```{r remedy002}
dat %>% count(color_1)


```
But this only tells us the first color selected, not how many times a color was selected. What if we want to count all the instances where "Moldy Book" was selected, across columns? Or getting a more succinct answer for all colors? Because these are not ordered in any way, and the respondent wasn't asked for an ordered preference, we need to count across the variables.

We can use `tidyr` for a quick solution:

```{r remedy003}

library(tidyr)

dat %>%
  gather(key, value, na.rm = TRUE) %>%
  count(value)

```

Good thing we checked! It turns out that Sparkle and Moldy Book are basically just as popular as Blue! If we had stopped with just checking the first color picked, our inference for color preference would have been way off.


