---
title: Week 10 Solutions
author: Ian Adams
date: '2021-03-28'
slug: week-10
categories:
  - R
tags:
  - '6002'
linktitle: Week 10 Soultions
summary: ~
lastmod: '2021-3-29T09:22:38-07:00'
toc: true
type: docs
menu:
  example:
    name: Week 10 Solutions
    weight: 4
---

```{r remedy001, include=FALSE}
library("nycflights13")
library("tidyverse")

```

# So you want to be a data scientist

This week can feel like a bit of a doozy, as the difficulty really ramps up. Something to remember - most data scientists spend most of their time cleaning, transforming, and *tidying* their data. That's what these chapters and questions are all about. So I want to make sure you have the correct answers, but more importantly, I want you to know **you are not alone**, this stuff is hard! I've put the quote in front of you before, but this little bit of wisdom from Hadley Wickham, lead author of our class textbook, is important during times like we're having right now:

> “There is no way of going from knowing nothing about a subject to knowing something about a subject without going through a period of much frustration and suckiness. **Push through. You’ll suck less**.”

Now, onto the answers!

## Exercise 5.2.4

<div class="question">
> Why is `NA ^ 0` not missing? Why is `NA | TRUE` not missing?
> Why is `FALSE & NA` not missing? Can you figure out the general rule?
> (`NA * 0` is a tricky counterexample!)

</div>

<div class="answer">

```{r}
NA ^ 0
```

`NA ^ 0 == 1` since for all numeric values $x ^ 0 = 1$.

```{r}
NA | TRUE
```

`NA | TRUE` is `TRUE` because anything **or** `TRUE` is `TRUE`. 
If the missing value were `TRUE`, then `TRUE | TRUE == TRUE`,
and if the missing value was `FALSE`, then `FALSE | TRUE == TRUE`.

```{r}
NA & FALSE
```

The value of `NA & FALSE` is `FALSE` because anything **and** `FALSE` is always `FALSE`.
If the missing value were `TRUE`, then `TRUE & FALSE == FALSE`,
and if the missing value was `FALSE`, then `FALSE & FALSE == FALSE`.

```{r}
NA | FALSE
```

For `NA | FALSE`, the value is unknown since `TRUE | FALSE == TRUE`, but `FALSE | FALSE == FALSE`.

```{r}
NA & TRUE
```

For `NA & TRUE`, the value is unknown since `FALSE & TRUE== FALSE`, but `TRUE & TRUE == TRUE`.

```{r}
NA * 0
```

Since $x * 0 = 0$ for all finite numbers we might expect `NA * 0 == 0`, but that's not the case.
The reason that `NA * 0 != 0` is that $0 \times \infty$ and $0 \times -\infty$ are undefined.
R represents undefined results as `NaN`, which is an abbreviation of "[not a number](https://en.wikipedia.org/wiki/NaN)".

```{r}
Inf * 0
-Inf * 0
```

</div>

## Exercise 5.3.1

<div class="question">
> How could you use `arrange()` to sort all missing values to the start? (Hint: use `is.na()`).

</div>

<div class="answer">

The `arrange()` function puts `NA` values last.
```{r}
arrange(flights, dep_time) %>%
  tail()
```
Using `desc()` does not change that.
```{r}
arrange(flights, desc(dep_time))
```

To put `NA` values first, we can add an indicator of whether the column has a missing value.
Then we sort by the missing indicator column and the column of interest. 
For example, to sort the data frame by departure time (`dep_time`) in ascending order but `NA` values first, run the following.
```{r}
arrange(flights, desc(is.na(dep_time)), dep_time)
```
The `flights`  will first be sorted by `desc(is.na(dep_time))`.
Since `desc(is.na(dep_time))` is either `TRUE` when `dep_time` is missing, or `FALSE`, when it is not, the rows with missing values of `dep_time` will come first, since `TRUE > FALSE`.

</div>

## Exercise 5.4.1

<div class="question">
> Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay` from flights.

</div>

<div class="answer">

These are a few ways to select columns.

-   Specify columns names as unquoted variable names.
    ```{r}
    select(flights, dep_time, dep_delay, arr_time, arr_delay)
    ```

-   Specify column names as strings.
    ```{r}
    select(flights, "dep_time", "dep_delay", "arr_time", "arr_delay")
    ```

-   Specify the column numbers of the variables.
    ```{r}
    select(flights, 4, 6, 7, 9)
    ```
    This works, but is not good practice for two reasons.
    First, the column location of variables may change, resulting in code that 
    may continue to run without error, but produce the wrong answer. 
    Second code is obfuscated, since it is not clear from the code which 
    variables are being selected. What variable does column 6 correspond to? 
    I just wrote the code, and I've already forgotten.

-   Specify the names of the variables with character vector and `any_of()` or `all_of()` 
    ```{r}
    select(flights, all_of(c("dep_time", "dep_delay", "arr_time", "arr_delay")))
    ```
    ```{r}
    select(flights, any_of(c("dep_time", "dep_delay", "arr_time", "arr_delay")))
    ```
    This is useful because the names of the variables can be stored in a 
    variable and passed to `all_of()` or `any_of()`.
    ```{r}
    variables <- c("dep_time", "dep_delay", "arr_time", "arr_delay")
    select(flights, all_of(variables))
    ```
    These two functions replace the deprecated function `one_of()`.

-   Selecting the variables by matching the start of their names using `starts_with()`.
    ```{r}
    select(flights, starts_with("dep_"), starts_with("arr_"))
    ```

-   Selecting the variables using regular expressions with `matches()`.
    Regular expressions provide a flexible way to match string patterns
    and are discussed in the [Strings](https://r4ds.had.co.nz/strings.html) chapter.
    ```{r}
    select(flights, matches("^(dep|arr)_(time|delay)$"))
    ```

-   Specify the names of the variables with a character vector and use the bang-bang operator (`!!`). 
    ```{r}
    variables <- c("dep_time", "dep_delay", "arr_time", "arr_delay")
    select(flights, !!variables)
    ```
    This and the following answers use the features of **tidy evaluation** not covered in R4DS but covered in the [Programming with dplyr](https://dplyr.tidyverse.org/articles/programming.html) vignette.

-   Specify the names of the variables in a character or list vector and use the bang-bang-bang operator.
    ```{r}
    variables <- c("dep_time", "dep_delay", "arr_time", "arr_delay")
    select(flights, !!!variables)
    ```  

-   Specify the unquoted names of the variables in a list using `syms()` and use the bang-bang-bang operator.
    ```{r}
    variables <- syms(c("dep_time", "dep_delay", "arr_time", "arr_delay"))
    select(flights, !!!variables)
    ```        

Some things that **don't** work are:

-   Matching the ends of their names using `ends_with()` since this will incorrectly
    include other variables. For example,
    ```{r}
    select(flights, ends_with("arr_time"), ends_with("dep_time"))
    ```

-   Matching the names using `contains()` since there is not a pattern that can
    include all these variables without incorrectly including others.
    ```{r}
    select(flights, contains("_time"), contains("arr_"))
    ```

</div>

## Exercise 5.5.2

<div class="question">

> Compare `air_time` with `arr_time - dep_time`. 
> What do you expect to see? 
> What do you see? 
> What do you need to do to fix it?


</div>

<div class="answer">

I expect that `air_time` is the difference between the arrival (`arr_time`) and departure times (`dep_time`).
In other words, `air_time = arr_time - dep_time`.

To check that this relationship, I'll first need to convert the times to a form more amenable to arithmetic operations using the same calculations as the [previous exercise](#exercise-5.5.1).
```{r}
flights_airtime <-
  mutate(flights,
    dep_time = (dep_time %/% 100 * 60 + dep_time %% 100) %% 1440,
    arr_time = (arr_time %/% 100 * 60 + arr_time %% 100) %% 1440,
    air_time_diff = air_time - arr_time + dep_time
  )
```

So, does `air_time = arr_time - dep_time`?
If so, there should be no flights with non-zero values of `air_time_diff`.
```{r}
nrow(filter(flights_airtime, air_time_diff != 0))
```

It turns out that there are many flights for which `air_time != arr_time - dep_time`.
Other than data errors, I can think of two reasons why `air_time` would not equal `arr_time - dep_time`.

1.  The flight passes midnight, so `arr_time < dep_time`.
    In these cases, the difference in airtime should be by 24 hours (1,440 minutes).

1.  The flight crosses time zones, and the total air time will be off by hours (multiples of 60). 
    All flights in `flights` departed from New York City and are domestic flights in the US.
    This means that flights will all be to the same or more westerly time zones.
    Given the time-zones in the US, the differences due to time-zone should be 60 minutes (Central)
    120 minutes (Mountain), 180 minutes (Pacific), 240 minutes (Alaska), or 300 minutes (Hawaii).
    
Both of these explanations have clear patterns that I would expect to see if they 
were true. 
In particular, in both cases, since time-zones and crossing midnight only affects the hour part of the time, all values of `air_time_diff` should be divisible by 60.
I'll visually check this hypothesis by plotting the distribution of `air_time_diff`.
If those two explanations are correct, distribution of `air_time_diff` should comprise only spikes at multiples of 60.
```{r}
ggplot(flights_airtime, aes(x = air_time_diff)) +
  geom_histogram(binwidth = 1)
```
This is not the case.
While, the distribution of `air_time_diff` has modes at multiples of 60 as hypothesized, 
it shows that there are many flights in which the difference between air time and local arrival and departure times is not divisible by 60.

Let's also look at flights with Los Angeles as a destination.
The discrepancy should be 180 minutes.
```{r}
ggplot(filter(flights_airtime, dest == "LAX"), aes(x = air_time_diff)) +
  geom_histogram(binwidth = 1)
```

To fix these time-zone issues, I would want to convert all the times to a date-time to handle overnight flights, and from local time to a common time zone, most likely [UTC](https://en.wikipedia.org/wiki/Coordinated_Universal_Time), to handle flights crossing time-zones.
The `tzone` column of `nycflights13::airports` gives the time-zone of each airport.
See the ["Dates and Times"](https://r4ds.had.co.nz/dates-and-times.html) for an introduction on working with date and time data.

But that still leaves the other differences unexplained. 
So what else might be going on? 
There seem to be too many problems for this to be data entry problems, so I'm probably missing something. 
So, I'll reread the documentation to make sure that I understand the definitions of `arr_time`, `dep_time`, and
`air_time`. 
The documentation contains a link to the source of the `flights` data, <https://www.transtats.bts.gov/DL_SelectFields.asp?Table_ID=236>.
This documentation shows that the `flights` data does not contain the variables `TaxiIn`, `TaxiOff`, `WheelsIn`, and `WheelsOff`.
It appears that the `air_time` variable refers to flight time, which is defined as the time between wheels-off (take-off) and wheels-in (landing).
But the flight time does not include time spent on the runway taxiing to and from gates.
With this new understanding of the data, I now know that the relationship between `air_time`, `arr_time`, and `dep_time` is `air_time <= arr_time - dep_time`, supposing that the time zones of `arr_time` and `dep_time` are in the same time zone.

</div>

## Exercise 5.6.7

> Brainstorm at least 5 different ways to assess the typical delay characteristics of a group of flights. 
> Consider the following scenarios:
> 
> -   A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of the time.
> -   A flight is always 10 minutes late.
> -   A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of the time.
> -   99% of the time a flight is on time. 1% of the time it’s 2 hours late.
> 
> Which is more important: arrival delay or departure delay?


</div>

<div class="answer">

What this question gets at is a fundamental question of data analysis: the cost function.
As analysts, the reason we are interested in flight delay because it is costly to passengers.
But it is worth thinking carefully about how it is costly and use that information in ranking and measuring these scenarios.

In many scenarios, arrival delay is more important.
In most cases, being arriving late is more costly to the passenger since it could disrupt the next stages of their travel, such as connecting flights or scheduled meetings.  
If a departure is delayed without affecting the arrival time, this delay will not have those affects plans nor does it affect the total time spent traveling.
This delay could be beneficial, if less time is spent in the cramped confines of the airplane itself, or a negative, if that delayed time is still spent in the cramped confines of the airplane on the runway.

Variation in arrival time is worse than consistency.
If a flight is always 30 minutes late and that delay is known, then it is as if the arrival time is that delayed time.
The traveler could easily plan for this. 
But higher variation in flight times makes it harder to plan.

### 5.6.7.2

<div class="question">

> Come up with another approach that will give you the same output as `not_cancelled %>% count(dest)` and `not_cancelled %>% count(tailnum, wt = distance)` (without using `count()`).

</div>

<div class="answer">

```{r not_cancelled}
not_cancelled <- flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay))
```

The first expression is the following.

```{r}
not_cancelled %>% 
  count(dest)
```

The `count()` function counts the number of instances within each group of variables.
Instead of using the `count()` function, we can combine the `group_by()` and `summarise()` verbs.

```{r}
not_cancelled %>%
  group_by(dest) %>%
  summarise(n = length(dest))
```

An alternative method for getting the number of observations in a data frame is the function `n()`.

```{r}
not_cancelled %>%
  group_by(dest) %>%
  summarise(n = n())
```

Another alternative to `count()` is to use `group_by()` followed by `tally()`.
In fact, `count()` is effectively a short-cut for `group_by()` followed by `tally()`.

```{r}
not_cancelled %>%
  group_by(tailnum) %>%
  tally()
```

The second expression also uses the `count()` function, but adds a `wt` argument.

```{r}
not_cancelled %>% 
  count(tailnum, wt = distance)
```

As before, we can replicate `count()` by combining the `group_by()` and `summarise()` verbs.
But this time instead of using `length()`, we will use `sum()` with the weighting variable.

```{r}
not_cancelled %>%
  group_by(tailnum) %>%
  summarise(n = sum(distance))
```

Like the previous example, we can also use the combination `group_by()` and `tally()`.
Any arguments to `tally()` are summed.

```{r}
not_cancelled %>%
  group_by(tailnum) %>%
  tally(distance)
```

</div>

### 5.6.7.3

<div class="question">
> Our definition of cancelled flights `(is.na(dep_delay) | is.na(arr_delay))` is slightly suboptimal. 
> Why? 
> Which is the most important column?

</div>

<div class="answer">

If a flight never departs, then it won't arrive.
A flight could also depart and not arrive if it crashes, or if it is redirected and lands in an airport other than its intended destination.
So the most important column is `arr_delay`, which indicates the amount of delay in arrival.
```{r}
filter(flights, !is.na(dep_delay), is.na(arr_delay)) %>%
  select(dep_time, arr_time, sched_arr_time, dep_delay, arr_delay)
```

In this data `dep_time` can be non-missing and `arr_delay` missing but `arr_time` not missing.
Some further [research](https://hyp.is/TsdRpofJEeqzs6-vUOfVBg/jrnold.github.io/r4ds-exercise-solutions/transform.html) found that these rows correspond to diverted flights.
The [BTS](https://www.transtats.bts.gov/DL_SelectFields.asp?Table_ID=236) database that is the source for the `flights` table contains additional information for diverted flights that is not included in the nycflights13 data. 
The source contains a column `DivArrDelay` with the description:

> Difference in minutes between scheduled and actual arrival time for a diverted flight reaching scheduled destination. 
> The `ArrDelay` column remains `NULL` for all diverted flights.

</div>

### 5.6.7.4

<div class="question">
> Look at the number of cancelled flights per day. 
> Is there a pattern?
> Is the proportion of cancelled flights related to the average delay?

</div>

<div class="answer">

One pattern in cancelled flights per day is that the number of cancelled flights increases with the total number of flights per day.
The proportion of cancelled flights increases with the average delay of flights.

To answer these questions, use definition of cancelled used in the 
chapter [Section 5.6.3](https://r4ds.had.co.nz/transform.html#counts) and the
relationship `!(is.na(arr_delay) & is.na(dep_delay))` is equal to 
`!is.na(arr_delay) | !is.na(dep_delay)` by [De Morgan's law](https://en.wikipedia.org/wiki/De_Morgan%27s_laws).

The first part of the question asks for any pattern in the number of cancelled flights per day.
I'll look at the relationship between the number of cancelled flights per day and the total number of flights in a day.
There should be an increasing relationship for two reasons.
First, if all flights are equally likely to be cancelled, then days with more flights should have a higher number of cancellations.
Second, it is likely that days with more flights would have a higher probability of cancellations because congestion itself can cause delays and any delay would affect more flights, and large delays can lead to cancellations.
```{r}
cancelled_per_day <- 
  flights %>%
  mutate(cancelled = (is.na(arr_delay) | is.na(dep_delay))) %>%
  group_by(year, month, day) %>%
  summarise(
    cancelled_num = sum(cancelled),
    flights_num = n(),
  )
```
Plotting `flights_num` against `cancelled_num` shows that the number of flights
cancelled increases with the total number of flights.
```{r}
ggplot(cancelled_per_day) +
  geom_point(aes(x = flights_num, y = cancelled_num)) 

```

The second part of the question asks whether there is a relationship between the proportion of flights cancelled and the average departure delay.
I implied this in my answer to the first part of the question, when I noted that increasing delays could result in increased cancellations.
The question does not specify which delay, so I will show the relationship for both.
```{r}
cancelled_and_delays <- 
  flights %>%
  mutate(cancelled = (is.na(arr_delay) | is.na(dep_delay))) %>%
  group_by(year, month, day) %>%
  summarise(
    cancelled_prop = mean(cancelled),
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    avg_arr_delay = mean(arr_delay, na.rm = TRUE)
  ) %>%
  ungroup()
```

There is a strong increasing relationship between both average departure delay and  
and average arrival delay and the proportion of cancelled flights.
```{r}
ggplot(cancelled_and_delays) +
  geom_point(aes(x = avg_dep_delay, y = cancelled_prop))

```

```{r}
ggplot(cancelled_and_delays) +
  geom_point(aes(x = avg_arr_delay, y = cancelled_prop))

```

</div>

### 5.6.7.5

<div class="question">
> Which carrier has the worst delays? 
> Challenge: can you disentangle the effects of bad airports vs. bad carriers? 
> Why/why not? 
> (Hint: think about `flights %>% group_by(carrier, dest) %>% summarise(n())`)

</div>

<div class="answer">

```{r}
flights %>%
  group_by(carrier) %>%
  summarise(arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(desc(arr_delay))
```

What airline corresponds to the `"F9"` carrier code?
```{r}
filter(airlines, carrier == "F9")
```

You can get part of the way to disentangling the effects of airports versus bad carriers by comparing the average delay of each carrier to the average delay of flights within a route (flights from the same origin to the same destination).
Comparing delays between carriers and within each route disentangles the effect of carriers and airports.
A better analysis would compare the average delay of a carrier's flights to the average delay of *all other* carrier's flights within a route.

```{r}
flights %>%
  filter(!is.na(arr_delay)) %>%
  # Total delay by carrier within each origin, dest
  group_by(origin, dest, carrier) %>%
  summarise(
    arr_delay = sum(arr_delay),
    flights = n()
  ) %>%
  # Total delay within each origin dest
  group_by(origin, dest) %>%
  mutate(
    arr_delay_total = sum(arr_delay),
    flights_total = sum(flights)
  ) %>%
  # average delay of each carrier - average delay of other carriers
  ungroup() %>%
  mutate(
    arr_delay_others = (arr_delay_total - arr_delay) /
      (flights_total - flights),
    arr_delay_mean = arr_delay / flights,
    arr_delay_diff = arr_delay_mean - arr_delay_others
  ) %>%
  # remove NaN values (when there is only one carrier)
  filter(is.finite(arr_delay_diff)) %>%
  # average over all airports it flies to
  group_by(carrier) %>%
  summarise(arr_delay_diff = mean(arr_delay_diff)) %>%
  arrange(desc(arr_delay_diff))
```

There are more sophisticated ways to do this analysis, however comparing the delay of flights within each route goes a long ways toward disentangling airport and carrier effects.
To see a more complete example of this analysis, see this FiveThirtyEight [piece](https://fivethirtyeight.com/features/the-best-and-worst-airlines-airports-and-flights-summer-2015-update/).

</div>

### 5.6.7.6

<div class="question">

> What does the sort argument to `count()` do?
> When might you use it?


</div>

<div class="answer">

The sort argument to `count()` sorts the results in order of `n`.
You could use this anytime you would run `count()` followed by `arrange()`.

For example, the following expression counts the number of flights to a destination and sorts the returned data from highest to lowest.
```{r}
flights %>%
  count(dest, sort = TRUE)
```

</div>

## Exercise 5.7.1

<div class="question">

> Refer back to the lists of useful mutate and filtering functions.
> Describe how each operation changes when you combine it with grouping.


</div>

<div class="answer">

Summary functions (`mean()`), offset functions (`lead()`, `lag()`), ranking functions (`min_rank()`, `row_number()`), operate within each group when used with `group_by()` in 
`mutate()` or `filter()`.
Arithmetic operators (`+`, `-`), logical operators (`<`, `==`), modular arithmetic operators (`%%`, `%/%`), logarithmic functions (`log`) are not affected by `group_by`.

Summary functions like `mean()`, `median()`, `sum()`, `std()` and others covered
in the section [Useful Summary Functions](https://r4ds.had.co.nz/transform.html#summarise-funs) 
calculate their values within each group when used with `mutate()` or `filter()` and `group_by()`.
```{r}
tibble(x = 1:9,
       group = rep(c("a", "b", "c"), each = 3)) %>%
  mutate(x_mean = mean(x)) %>%
  group_by(group) %>%
  mutate(x_mean_2 = mean(x))
```

Arithmetic operators `+`, `-`, `*`, `/`, `^` are not affected by `group_by()`.
```{r}
tibble(x = 1:9,
       group = rep(c("a", "b", "c"), each = 3)) %>%
  mutate(y = x + 2) %>%
  group_by(group) %>%
  mutate(z = x + 2)
```

The modular arithmetic operators `%/%` and `%%` are not affected by `group_by()`
```{r}
tibble(x = 1:9,
       group = rep(c("a", "b", "c"), each = 3)) %>%
  mutate(y = x %% 2) %>%
  group_by(group) %>%
  mutate(z = x %% 2)
```

The logarithmic functions `log()`, `log2()`, and `log10()` are not affected by
`group_by()`.
```{r}
tibble(x = 1:9,
       group = rep(c("a", "b", "c"), each = 3)) %>%
  mutate(y = log(x)) %>%
  group_by(group) %>%
  mutate(z = log(x))
```

The offset functions `lead()` and `lag()` respect the groupings in `group_by()`.
The functions `lag()` and `lead()` will only return values within each group.
```{r}
tibble(x = 1:9,
       group = rep(c("a", "b", "c"), each = 3)) %>%
  group_by(group) %>%
  mutate(lag_x = lag(x),
         lead_x = lead(x))
```

The cumulative and rolling aggregate functions `cumsum()`, `cumprod()`, `cummin()`, `cummax()`, and `cummean()` calculate values within each group.
```{r}
tibble(x = 1:9,
       group = rep(c("a", "b", "c"), each = 3)) %>%
  mutate(x_cumsum = cumsum(x)) %>%
  group_by(group) %>%
  mutate(x_cumsum_2 = cumsum(x))

```

Logical comparisons, `<`, `<=`, `>`, `>=`, `!=`, and `==` are not affected by `group_by()`.
```{r}
tibble(x = 1:9,
       y = 9:1,
       group = rep(c("a", "b", "c"), each = 3)) %>%
  mutate(x_lte_y = x <= y) %>%
  group_by(group) %>%
  mutate(x_lte_y_2 = x <= y)
```

Ranking functions like `min_rank()` work within each group when used with `group_by()`.
```{r}
tibble(x = 1:9,
       group = rep(c("a", "b", "c"), each = 3)) %>%
  mutate(rnk = min_rank(x)) %>%
  group_by(group) %>%
  mutate(rnk2 = min_rank(x))
```

Though not asked in the question, note that `arrange()` ignores groups when sorting values.
```{r}
tibble(x = runif(9),
       group = rep(c("a", "b", "c"), each = 3)) %>%
  group_by(group) %>%
  arrange(x)

```
However, the order of values from `arrange()` can interact with groups when 
used with functions that rely on the ordering of elements, such as `lead()`, `lag()`,
or `cumsum()`.
```{r}
tibble(group = rep(c("a", "b", "c"), each = 3), 
       x = runif(9)) %>%
  group_by(group) %>%
  arrange(x) %>%
  mutate(lag_x = lag(x))

```

</div>

### 5.7.1.2

<div class="question">

> Which plane (`tailnum`) has the worst on-time record?

</div>

<div class="answer">

The question does not define a way to measure on-time record, so I will consider two metrics:

1.  proportion of flights not delayed or cancelled, and
1.  mean arrival delay.

The first metric is the proportion of not-cancelled and on-time flights.
I use the presence of an arrival time as an indicator that a flight was not cancelled.
However, there are many planes that have never flown an on-time flight.
Additionally, many of the planes that have the lowest proportion of on-time flights have only flown a small number of flights.
```{r}
flights %>%
  filter(!is.na(tailnum)) %>%
  mutate(on_time = !is.na(arr_time) & (arr_delay <= 0)) %>%
  group_by(tailnum) %>%
  summarise(on_time = mean(on_time), n = n()) %>%
  filter(min_rank(on_time) == 1)

```

So, I will remove planes that flew at least 20 flights.
The choice of 20 was chosen because it round number near the first quartile of the number of flights by plane.[^delay][^count]
```{r}
quantile(count(flights, tailnum)$n)
```

The plane with the worst on time record that flew at least 20 flights is:
```{r}
flights %>%
  filter(!is.na(tailnum), is.na(arr_time) | !is.na(arr_delay)) %>%
  mutate(on_time = !is.na(arr_time) & (arr_delay <= 0)) %>%
  group_by(tailnum) %>%
  summarise(on_time = mean(on_time), n = n()) %>%
  filter(n >= 20) %>%
  filter(min_rank(on_time) == 1)

```

There are cases where `arr_delay` is missing but `arr_time` is not missing.
I have not debugged the cause of this bad data, so these rows are dropped for
the purposes of this exercise.

The second metric is the mean minutes delayed.
As with the previous metric, I will only consider planes which flew least 20 flights.
A different plane has the worst on-time record when measured as average minutes delayed.
```{r}
flights %>%
  filter(!is.na(arr_delay)) %>%
  group_by(tailnum) %>%
  summarise(arr_delay = mean(arr_delay), n = n()) %>%
  filter(n >= 20) %>%
  filter(min_rank(desc(arr_delay)) == 1)
```

</div>

### 5.7.1.3

<div class="question">

> What time of day should you fly if you want to avoid delays as much as possible?

</div>

<div class="answer">

Let's group by the hour of the flight.
The earlier the flight is scheduled, the lower its expected delay.
This is intuitive as delays will affect later flights. 
Morning flights have fewer (if any) previous flights that can delay them.

```{r}
flights %>%
  group_by(hour) %>%
  summarise(arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(arr_delay)
```

</div>

### 5.7.1.4

<div class="question">

> For each destination, compute the total minutes of delay. 
> For each flight, compute the proportion of the total delay for its destination.

</div>

<div class="answer">

The key to answering this question is to only include delayed flights when calculating the total delay and proportion of delay.

```{r}
flights %>%
  filter(arr_delay > 0) %>%
  group_by(dest) %>%
  mutate(
    arr_delay_total = sum(arr_delay),
    arr_delay_prop = arr_delay / arr_delay_total
  ) %>%
  select(dest, month, day, dep_time, carrier, flight,
         arr_delay, arr_delay_prop) %>%
  arrange(dest, desc(arr_delay_prop))
```

There is some ambiguity in the meaning of the term *flights* in the question.
The first example defined a flight as a row in the `flights` table, which is a trip by an aircraft from an airport at a particular date and time. 
However, *flight* could also refer to the [flight number](https://en.wikipedia.org/wiki/Flight_number), which is the code a carrier uses for an airline service of a route.
For example, `AA1` is the flight number of the 09:00 American Airlines flight between JFK and LAX.
The flight number is contained in the `flights$flight` column, though what is called a "flight" is a combination of the `flights$carrier` and `flights$flight` columns.

```{r}
flights %>%
  filter(arr_delay > 0) %>%
  group_by(dest, origin, carrier, flight) %>%
  summarise(arr_delay = sum(arr_delay)) %>%
  group_by(dest) %>%
  mutate(
    arr_delay_prop = arr_delay / sum(arr_delay)
  ) %>%
  arrange(dest, desc(arr_delay_prop)) %>%
  select(carrier, flight, origin, dest, arr_delay_prop)
```

</div>

### 5.7.1.5

<div class="question">

>Delays are typically temporally correlated: even once the problem that caused the initial delay has been resolved, later >flights are delayed to allow earlier flights to leave. Using `lag()` explore how the delay of a flight is related to the >delay of the immediately preceding flight.

</div>

<div class="answer">

This calculates the departure delay of the preceding flight from the same airport.
```{r}
lagged_delays <- flights %>%
  arrange(origin, month, day, dep_time) %>%
  group_by(origin) %>%
  mutate(dep_delay_lag = lag(dep_delay)) %>%
  filter(!is.na(dep_delay), !is.na(dep_delay_lag))
```

This plots the relationship between the mean delay of a flight for all values of the previous flight.
For delays less than two hours, the relationship between the delay of the preceding flight and the current flight is nearly a line.
After that the relationship becomes more variable, as long-delayed flights are interspersed with flights leaving on-time.
After about 8-hours, a delayed flight is likely to be followed by a flight leaving on time.

```{r message=FALSE}
lagged_delays %>%
  group_by(dep_delay_lag) %>%
  summarise(dep_delay_mean = mean(dep_delay)) %>%
  ggplot(aes(y = dep_delay_mean, x = dep_delay_lag)) +
  geom_point() +
  scale_x_continuous(breaks = seq(0, 1500, by = 120)) +
  labs(y = "Departure Delay", x = "Previous Departure Delay")
```

The overall relationship looks similar in all three origin airports.
```{r}
lagged_delays %>%
  group_by(origin, dep_delay_lag) %>%
  summarise(dep_delay_mean = mean(dep_delay)) %>%
  ggplot(aes(y = dep_delay_mean, x = dep_delay_lag)) +
  geom_point() +
  facet_wrap(~ origin, ncol=1) +
  labs(y = "Departure Delay", x = "Previous Departure Delay")
```

</div>

### 5.7.1.6

<div class="question">

> Look at each destination. Can you find flights that are suspiciously fast? 
(i.e. flights that represent a potential data entry error).
Compute the air time of a flight relative to the shortest flight to that destination.
Which flights were most delayed in the air?

</div>

<div class="answer">

When calculating this answer we should only compare flights within the same (origin, destination) pair.

To find unusual observations, we need to first put them on the same scale.
I will [standardize](https://en.wikipedia.org/wiki/Standard_score)
values by subtracting the mean from each and then dividing each by the standard deviation.
$$
\mathsf{standardized}(x) = \frac{x - \mathsf{mean}(x)}{\mathsf{sd}(x)} .
$$
A standardized variable is often called a $z$-score.
The units of the standardized variable are standard deviations from the mean.
This will put the flight times from different routes on the same scale.
The larger the magnitude of the standardized variable for an observation, the more unusual the observation is.
Flights with negative values of the standardized variable are faster than the 
mean flight for that route, while those with positive values are slower than
the mean flight for that route.

```{r}
standardized_flights <- flights %>%
  filter(!is.na(air_time)) %>%
  group_by(dest, origin) %>%
  mutate(
    air_time_mean = mean(air_time),
    air_time_sd = sd(air_time),
    n = n()
  ) %>%
  ungroup() %>%
  mutate(air_time_standard = (air_time - air_time_mean) / (air_time_sd + 1))
```
I add 1 to the denominator and numerator to avoid dividing by zero.
Note that the `ungroup()` here is not necessary. However, I will be using 
this data frame later. Through experience, I have found that I have fewer bugs
when I keep a data frame grouped for only those verbs that need it.
If I did not `ungroup()` this data frame, the `arrange()` used later would 
not work as expected.  It is better to err on the side of using `ungroup()`
when unnecessary.

The distribution of the standardized air flights has long right tail.
```{r}
ggplot(standardized_flights, aes(x = air_time_standard)) +
  geom_density()
```

Unusually fast flights are those flights with the smallest standardized values.
```{r}
standardized_flights %>%
  arrange(air_time_standard) %>%
  select(
    carrier, flight, origin, dest, month, day,
    air_time, air_time_mean, air_time_standard
  ) %>%
  head(10) %>%
  print(width = Inf)
```

I used `width = Inf` to ensure that all columns will be printed.

```{r include=FALSE,purl=FALSE}
format_ymd <- function(y, m, d) {
  format(lubridate::make_date(y, m, d), "%Y-%m-%d")
}
format_time <- function(x) {
  format(
    lubridate::make_datetime(hour = x %/% 100, min = x %% 100),
    "%H:%M"
  )
}
fastest_flight <- standardized_flights %>%
  arrange(air_time_standard) %>%
  slice(1L) %>%
  mutate(
    date = format_ymd(year, month, day),
    time = format_time(dep_time),
    flightnum = str_c(carrier, flight)
  )
```
The fastest flight is `r fastest_flight$flightnum` from `r fastest_flight$origin` to
`r fastest_flight$dest` which departed on 
`r fastest_flight$date` at `r fastest_flight$time`.
It has an air time of `r round(fastest_flight$air_time)` minutes, compared to an average
flight time of `r round(fastest_flight$air_time_mean)` minutes for its route.
This is `r round(abs(fastest_flight$air_time_standard), 1)` standard deviations below
the average flight on its route.

It is important to note that this does not necessarily imply that there was a data entry error. 
We should check these flights to see whether there was some reason for the difference.
It may be that we are missing some piece of information that explains these unusual times.

A potential issue with the way that we standardized the flights is that the mean and standard deviation used to calculate are sensitive to outliers and outliers is what we are looking for.
Instead of standardizing variables with the mean and variance, we could use the median
as a measure of central tendency and the interquartile range (IQR) as a measure of spread.
The median and IQR are more [resistant to outliers](https://en.wikipedia.org/wiki/Robust_statistics) than the mean and standard deviation.
The following method uses the median and inter-quartile range, which are less sensitive to outliers.

```{r}
standardized_flights2 <- flights %>%
  filter(!is.na(air_time)) %>%
  group_by(dest, origin) %>%
  mutate(
    air_time_median = median(air_time),
    air_time_iqr = IQR(air_time),
    n = n(),
    air_time_standard = (air_time - air_time_median) / air_time_iqr)
```

The distribution of the standardized air flights using this new definition
also has long right tail of slow flights.
```{r}
ggplot(standardized_flights2, aes(x = air_time_standard)) +
  geom_density()
```

Unusually fast flights are those flights with the smallest standardized values.
```{r}
standardized_flights2 %>%
  arrange(air_time_standard) %>%
  select(
    carrier, flight, origin, dest, month, day, air_time,
    air_time_median, air_time_standard
  ) %>%
  head(10) %>%
  print(width = Inf)
```

All of these answers have relied only on using a distribution of comparable observations to find unusual observations. 
In this case, the comparable observations were flights from the same origin to the same destination.
Apart from our knowledge that flights from the same origin to the same destination should have similar air times, we have not used any other domain-specific knowledge.
But we know much more about this problem.
The most obvious piece of knowledge we have is that we know that flights cannot travel back in time, so there should never be a flight with a negative airtime.
But we also know that aircraft have maximum speeds.
While different aircraft have different [cruising speeds](https://en.wikipedia.org/wiki/Cruise_(aeronautics)), commercial airliners
typically cruise at air speeds around 547–575 mph.
Calculating the ground speed of aircraft is complicated by the way in which winds, especially the influence of wind, especially jet streams, on the ground-speed of flights.
A strong tailwind can increase ground-speed of the aircraft by [200 mph](https://www.wired.com/story/norwegian-air-transatlantic-speed-record/).
Apart from the retired [Concorde](https://en.wikipedia.org/wiki/Concorde).
For example, in 2018, [a transatlantic flight](https://www.wired.com/story/norwegian-air-transatlantic-speed-record/) 
traveled at 770 mph due to a strong jet stream tailwind.
This means that any flight traveling at speeds greater than 800 mph is implausible, 
and it may be worth checking flights traveling at greater than 600 or 700 mph.
Ground speed could also be used to identify aircraft flying implausibly slow.
Joining flights data with the air craft type in the `planes` table and getting
information about typical or top speeds of those aircraft could provide a more 
detailed way to identify implausibly fast or slow flights.
Additional data on high altitude wind speeds at the time of the flight would further help.

Knowing the substance of the data analysis at hand is one of the most important 
tools of a data scientist. The tools of statistics are a complement, not a 
substitute, for that knowledge.

With that in mind, Let's plot the distribution of the ground speed of flights. 
The modal flight in this data has a ground speed of between 400 and 500 mph.
The distribution of ground speeds has a large left tail of slower flights below
400 mph constituting the majority.
There are very few flights with a ground speed over 500 mph.

```{r}
flights %>%
  mutate(mph = distance / (air_time / 60)) %>%
  ggplot(aes(x = mph)) +
  geom_histogram(binwidth = 10)
```

```{r include=FALSE,purl=FALSE}
flights_mph <- flights %>%
  mutate(mph = distance / (air_time / 60))
fastest_flight_mph <- arrange(ungroup(flights_mph), desc(mph)) %>%
  slice(1L)
over_600mph <- filter(flights_mph, mph > 600) %>% nrow()
```

The fastest flight is the same one identified as the largest outlier earlier.
Its ground speed was `r round(fastest_flight_mph$mph)` mph. 
This is fast for a commercial jet, but not impossible. 

```{r}
flights %>%
  mutate(mph = distance / (air_time / 60)) %>%
  arrange(desc(mph)) %>%
  select(mph, flight, carrier, flight, month, day, dep_time) %>%
  head(5)
```

One explanation for unusually fast flights is that they are "making up time" in the air by flying faster.
Commercial aircraft do not fly at their top speed since the airlines are also concerned about fuel consumption.
But, if a flight is delayed on the ground, it may fly faster than usual in order to avoid a late arrival.
So, I would expect that some of the unusually fast flights were delayed on departure.

```{r}
flights %>%
  mutate(mph = distance / (air_time / 60)) %>%
  arrange(desc(mph)) %>%
  select(
    origin, dest, mph, year, month, day, dep_time, flight, carrier,
    dep_delay, arr_delay
  )
head(5)
```

Five of the top ten flights had departure delays, and three of those were
able to make up that time in the air and arrive ahead of schedule.

Overall, there were a few flights that seemed unusually fast, but they all 
fall into the realm of plausibility and likely are not data entry problems.
[Ed. Please correct me if I am missing something]

<!--
Similarly, the longest [regularly scheduled flight](https://en.wikipedia.org/wiki/Longest_flights#Record_flights) is Newark
to Signapore with a duration of 18 hours 30--45 minutes.
Thus, we should never observe any commercial flight longer than 19 hours in our
data.
-->

The second part of the question asks us to compare flights to the fastest flight
on a route to find the flights most delayed in the air. I will calculate the 
amount a flight is delayed in air in two ways. 
The first is the absolute delay, defined as the number of minutes longer than the fastest flight on that route,`air_time - min(air_time)`. 
The second is the relative delay, which is the percentage increase in air time relative to the time of the fastest flight
along that route, `(air_time - min(air_time)) / min(air_time) * 100`.

```{r}
air_time_delayed <-
  flights %>%
  group_by(origin, dest) %>%
  mutate(
    air_time_min = min(air_time, na.rm = TRUE),
    air_time_delay = air_time - air_time_min,
    air_time_delay_pct = air_time_delay / air_time_min * 100
  )
```
```{r include=FALSE,purl=FALSE}
most_delayed <- arrange(
  air_time_delayed,
  desc(air_time_delay)
) %>%
  mutate(
    date = format_ymd(year, month, day),
    time = format_time(dep_time),
    carrier_flight = str_c(carrier, flight)
  ) %>%
  head(1L)

most_delayed_pct <- arrange(air_time_delayed, desc(air_time_delay_pct)) %>%
  mutate(
    date = format_ymd(year, month, day),
    time = format_time(dep_time),
    carrier_flight = str_c(carrier, flight)
  ) %>%
  head(1L)
```

The most delayed flight in air in minutes was `r most_delayed$carrier_flight` 
from `r most_delayed$origin` to `r most_delayed$dest` which departed on 
`r most_delayed$date` at `r most_delayed$time`. It took
`r most_delayed$air_time_delay` minutes longer than the flight with the shortest
air time on its route.

```{r}
air_time_delayed %>%
  arrange(desc(air_time_delay)) %>%
  select(
    air_time_delay, carrier, flight,
    origin, dest, year, month, day, dep_time,
    air_time, air_time_min
  ) %>%
  head() %>%
  print(width = Inf)
```

The most delayed flight in air as a percentage of the fastest flight along that
route was `r most_delayed_pct$carrier_flight` 
from `r most_delayed_pct$origin` to `r most_delayed_pct$dest` departing on `r most_delayed_pct$date` at `r most_delayed_pct$time`.
It took `r round(most_delayed_pct$air_time_delay_pct)`% longer than the 
flight with the shortest air time on its route.

```{r}
air_time_delayed %>%
  arrange(desc(air_time_delay)) %>%
  select(
    air_time_delay_pct, carrier, flight,
    origin, dest, year, month, day, dep_time,
    air_time, air_time_min
  ) %>%
  head() %>%
  print(width = Inf)
```

</div>

### 5.7.1.7

<div class="question">

> Find all destinations that are flown by at least two carriers. 
> Use that information to rank the carriers.

</div>

<div class="answer">

To restate this question, we are asked to rank airlines by the number of destinations that they fly to, considering only those airports that are flown to by two or more airlines.
There are two steps to calculating this ranking.
First, find all airports serviced by two or more carriers.
Then, rank carriers by the number of those destinations that they service.

```{r}
flights %>%
   # find all airports with > 1 carrier
   group_by(dest) %>%
   mutate(n_carriers = n_distinct(carrier)) %>%
   filter(n_carriers > 1) %>%
   # rank carriers by numer of destinations
   group_by(carrier) %>%
   summarize(n_dest = n_distinct(dest)) %>%
   arrange(desc(n_dest))
```

The carrier `"EV"` flies to the most destinations, considering only airports flown to by two or more carriers. What airline does the `"EV"` carrier code correspond to?
```{r}
filter(airlines, carrier == "EV")
```
Unless you know the airplane industry, it is likely that you don't recognize [ExpressJet](https://en.wikipedia.org/wiki/ExpressJet); I certainly didn't.
It is a regional airline that partners with major airlines to fly from hubs (larger airports) to smaller airports.
This means that many of the shorter flights of major carriers are operated by ExpressJet.
This business model explains why ExpressJet services the most destinations.

Among the airlines that fly to only one destination from New York are Alaska Airlines
and Hawaiian Airlines.
```{r}
filter(airlines, carrier %in% c("AS", "F9", "HA"))
```

</div>


### 5.7.1.8

<div class="question">
For each plane, count the number of flights before the first delay of greater than 1 hour.
</div>

<div class="answer">

The question does not specify arrival or departure delay.
I consider `dep_delay` in this answer, though similar code could be used for `arr_delay`.

```{r}
flights %>%
  # sort in increasing order
  select(tailnum, year, month,day, dep_delay) %>%
  filter(!is.na(dep_delay)) %>%
  arrange(tailnum, year, month, day) %>%
  group_by(tailnum) %>%
  # cumulative number of flights delayed over one hour
  mutate(cumulative_hr_delays = cumsum(dep_delay > 60)) %>%
  # count the number of flights == 0
  summarise(total_flights = sum(cumulative_hr_delays < 1)) %>%
  arrange(total_flights)
```

</div>

[^delay]: We could address this issue using a statistical model, but that is outside
          the scope of this text.

[^count]: The `count()` function is introduced in [Chapter 5.6](https://r4ds.had.co.nz/transform.html#counts). It returns the count of 
          rows by group. In this case, the number of rows in `flights` for each
          `tailnum`. The data frame that `count()` returns has columns for the 
          groups, and a column `n`, which contains that count.
