---
title: "Analyzing Police Violence in America: Updated Data Through 2025"
subtitle: 'Static Visualizations from Mapping Police Violence Database'
date: '2026-01-04'
authors: []
tags: [police, data visualization, R]
categories: [R, Police, Data Analysis]
summary: 'An automatically-updating analysis of police killings in the United States using data from Mapping Police Violence, with static visualizations showing temporal trends, demographic patterns, and geographic distributions.'
lastmod: '`r Sys.Date()`'
featured: yes
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7,
  dpi = 300
)
```

## Important Data Limitations

**This analysis presents data on police killings, which represents only a subset of all police uses of lethal force.** The data captures fatal encounters but does not include:

- **Non-fatal police shootings** (research suggests that a slight majority of police shootings are survived)
- Other uses of lethal force that did not result in death
- Uses of force that did not involve lethal weapons

Therefore, **these numbers undercount the total instances of lethal force used by police.** When interpreting these statistics, it's critical to remember that we are looking at deaths only, not all incidents where officers used potentially lethal force.

Additionally, this data represents **descriptive statistics only** and cannot establish causal relationships without additional controls for local crime rates, police deployment patterns, population demographics, socioeconomic factors, and jurisdictional policies.

This is a **vibe-coded prototype**: a rapid, exploratory build to see how far one can get in standing up a public-data dashboard with automated ingestion and standard visual analytics.

**Critical disclaimer:** I have not independently verified the MPV data, nor have I formally audited or validated my cleaning and harmonization routines. Treat everything here as exploratory and illustrative. The outputs are not suitable for professional, policy, legal, journalistic, or operational use, and should not be relied upon as an authoritative accounting of levels, rates, rankings, or trends.

---

## Automatic Data Updates

This post automatically downloads the latest data from [Mapping Police Violence](https://mappingpoliceviolence.us/) whenever the website is rebuilt. The visualizations below will update as new data becomes available.

```{r load-packages}
library(tidyverse)
library(readr)
library(readxl)
library(lubridate)
library(gghighlight)
library(ggthemes)
library(ggrepel)
library(slider)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

# Suppress package startup messages
suppressPackageStartupMessages({
  library(tidyverse)
})
```

```{r download-data, cache=FALSE}
# Download MPV data
mpv_url <- "https://mappingpoliceviolence.us/s/MPVDatasetDownload.xlsx"
destfile <- tempfile(fileext = ".xlsx")

# Download fresh data each time
download.file(mpv_url, destfile, mode = "wb", quiet = TRUE)

# Read the Excel file
df_mpv <- read_excel(destfile, guess_max = 10000)
```

```{r clean-data}
# Clean column names
names(df_mpv) <- names(df_mpv) %>%
  str_replace_all(" ", "_") %>%
  str_replace_all("[^[:alnum:]_]", "") %>%
  str_to_lower()

# Find the date column
date_col <- names(df_mpv) %>% str_subset("date") %>% first()

# Parse date field - ALL deaths
df_all <- df_mpv %>%
  rename(date = all_of(date_col)) %>%
  mutate(
    date = as.Date(date),
    year = year(date),
    month = month(date, label = TRUE, abbr = TRUE),
    day = wday(date, label = TRUE, abbr = TRUE)
  ) %>%
  filter(!is.na(date))

# Create shootings-only dataset
df_shootings <- df_all %>%
  filter(str_detect(cause_of_death, "Gunshot"))

# Store key dates
current_date <- Sys.Date()
formatted_date <- format(current_date, "%B %d, %Y")
min_year <- min(df_all$year)
latest_year <- max(df_all$year)
```

## Data Summary

**Data current as of `r formatted_date`**

- **All Deaths:** `r format(nrow(df_all), big.mark=",")` incidents from `r format(min(df_all$date), "%B %d, %Y")` to `r format(max(df_all$date), "%B %d, %Y")`
- **Fatal Shootings:** `r format(nrow(df_shootings), big.mark=",")` incidents (`r round(100 * nrow(df_shootings) / nrow(df_all), 1)`%)
- **Years covered:** `r min_year` - `r latest_year`

---

## Cumulative Trends

These charts show the cumulative count of police killings throughout each year, allowing year-over-year comparison.

```{r cumulative-all-deaths, fig.cap="Cumulative count of all police killings by method (gunshot, taser, vehicle, restraint, etc.)"}
# Function to create cumulative plot
create_cumulative_plot <- function(data, title_suffix, y_limit, caption_note) {
  shootings_data <- data %>%
    arrange(date) %>%
    mutate(year = as.numeric(format(date, "%Y"))) %>%
    group_by(year, date) %>%
    summarize(daily_count = n(), .groups = 'drop') %>%
    ungroup() %>%
    group_by(year) %>%
    mutate(cum_count = cumsum(daily_count)) %>%
    ungroup()

  min_year <- min(shootings_data$year)
  latest_year <- max(shootings_data$year)
  plot_title <- paste("Cumulative U.S. Police", title_suffix, "by Year,", min_year, "-", latest_year)

  p <- ggplot(shootings_data, aes(x = format(date, "%m-%d"), y = cum_count, group = factor(year))) +
    geom_point(aes(color = factor(year)), alpha = 0.5, size = 1) +
    gghighlight(year == latest_year, use_direct_label = FALSE,
                unhighlighted_params = list(alpha = 0.3)) +
    scale_x_discrete(name = "Month",
                     breaks = format(seq(as.Date("2022-01-01"), as.Date("2022-12-31"), by = "1 month"), "%m-%d"),
                     labels = format(seq(as.Date("2022-01-01"), as.Date("2022-12-31"), by = "1 month"), "%b"),
                     expand = c(0.01, 0.9)) +
    scale_y_continuous(name = "Cumulative Count of Deaths", limits = c(0, y_limit)) +
    theme_tufte() +
    theme(panel.grid.minor = element_blank(),
          legend.position = "none",
          axis.text.x = element_text(angle = 90, vjust = 0.5),
          axis.text.y = element_text(angle = 0, hjust = 0.5),
          plot.title = element_text(size = 14, hjust = 0, face = "bold"),
          plot.title.position = "plot",
          plot.caption = element_text(size = 10, hjust = 1)) +
    geom_text_repel(data = shootings_data %>%
                      group_by(year) %>%
                      slice_max(date, n = 1) %>%
                      ungroup(),
                    aes(x = format(date, "%m-%d"), y = cum_count,
                        label = paste(year, "(", cum_count, ")")),
                    size = 3, hjust = 0, direction = "y",
                    color = "black", fontface = "bold") +
    labs(title = plot_title,
         subtitle = paste("Data through", format(max(data$date), "%B %d, %Y")),
         caption = paste("Source: Ian T. Adams | Data: Mapping Police Violence\n", caption_note))

  return(p)
}

# Create plot for all deaths
create_cumulative_plot(df_all, "Killings (All Methods)", 1300,
                       "Includes all causes: gunshot, taser, vehicle, restraint, etc.")
```

```{r cumulative-shootings, fig.cap="Cumulative count of fatal police shootings only"}
# Create plot for shootings only
create_cumulative_plot(df_shootings, "Fatal Shootings", 1150,
                       "Filtered to cases with 'Gunshot' in cause of death\nNote: This excludes non-fatal shootings")
```

---

## Demographic Patterns

### Race and Ethnicity

```{r race-cleanup}
# Create race variable
df_shootings <- df_shootings %>%
  mutate(race_clean = case_when(
    is.na(victims_race) | victims_race == "" | victims_race == "Unknown race" ~ "Unknown",
    str_detect(victims_race, "White") ~ "White",
    str_detect(victims_race, "Black") ~ "Black",
    str_detect(victims_race, "Hispanic") | str_detect(victims_race, "Latino") ~ "Hispanic",
    str_detect(victims_race, "Asian") ~ "Asian",
    str_detect(victims_race, "Native American") ~ "Native American",
    str_detect(victims_race, "Pacific Islander") ~ "Pacific Islander",
    TRUE ~ "Other"
  ))

# Reorder by frequency
df_shootings$race_clean <- factor(df_shootings$race_clean,
                                  levels = names(sort(table(df_shootings$race_clean),
                                                      decreasing = TRUE)))
```

```{r race-distribution, fig.cap="Distribution of fatal police shootings by race/ethnicity"}
plot_title <- paste("Distribution of Police Fatal Shootings by Race,", min_year, "-", latest_year)

ggplot(df_shootings, aes(x = race_clean)) +
  geom_bar(fill = "steelblue", color = "black") +
  geom_text(aes(label = sprintf("%0.1f%%", (..count..)/sum(..count..)*100)),
            stat = "count", vjust = -0.5, color = "black", size = 3.5) +
  theme_tufte() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 14, hjust = 0, face = "bold"),
        plot.title.position = "plot") +
  labs(y = "Count", title = plot_title,
       subtitle = paste("as of", formatted_date),
       caption = "Source: Ian T. Adams | Data: Mapping Police Violence")
```

```{r race-by-year, fig.width=12, fig.cap="Fatal police shootings by year and race, with percentage of unknown race highlighted"}
# Calculate unknown percentages
colorsb <- c("#0072b2","#d55e00", "#009e73", "#f0e442", "#56b4e9", "#cc79a7", "#999999", "#E69F00")

df_percent <- df_shootings %>%
  group_by(year, race_clean) %>%
  summarize(count = n(), .groups = 'drop') %>%
  group_by(year) %>%
  mutate(total = sum(count),
         percent = count / total * 100) %>%
  filter(race_clean == "Unknown")

# Create base plot
p_race_year <- df_shootings %>%
  ggplot(aes(x = year, fill = race_clean)) +
  geom_bar(col = 'white') +
  scale_fill_manual(values = colorsb) +
  labs(x = 'Year', y = 'Counts',
       title = 'Fatal Police Shootings by Year and Race',
       subtitle = paste0('Percentages show unknown race category (as of ', formatted_date, ')'),
       caption = "Source: Ian T. Adams | Data: Mapping Police Violence",
       fill = 'Race') +
  scale_x_continuous(breaks = seq(min(df_shootings$year), max(df_shootings$year), by = 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Add percentage labels
p_race_year +
  geom_text(data = df_percent,
            aes(x = year, y = count, label = paste0(round(percent, 1), "%")),
            position = position_stack(vjust = 0.4), color = "black", size = 3)
```

### Age Distribution

```{r age-analysis}
# Create numeric age variable
df_shootings <- df_shootings %>%
  mutate(age_numeric = as.numeric(victims_age))
```

```{r age-distribution, fig.cap="Age distribution of police shooting victims"}
ggplot(df_shootings %>% filter(!is.na(age_numeric)), aes(x = age_numeric)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(x = "Age", y = "Count",
       title = "Age Distribution of Police Shooting Victims",
       subtitle = paste(min_year, "-", latest_year, "| Mean age:",
                       round(mean(df_shootings$age_numeric, na.rm = TRUE), 1),
                       "| Median age:", median(df_shootings$age_numeric, na.rm = TRUE)),
       caption = "Source: Ian T. Adams | Data: Mapping Police Violence")
```

```{r age-race-distribution, fig.width=10, fig.cap="Age distribution by race for the three largest racial groups"}
age_race <- df_shootings %>%
  filter(!is.na(age_numeric) & race_clean %in% c("White", "Black", "Hispanic")) %>%
  mutate(age_group = cut(age_numeric,
                         breaks = c(0, 18, 25, 35, 45, 55, 65, 100),
                         labels = c("<18", "18-24", "25-34", "35-44", "45-54", "55-64", "65+")))

ggplot(age_race, aes(x = age_group, fill = race_clean)) +
  geom_bar(position = "dodge") +
  scale_fill_brewer(palette = "Set1", name = "Race") +
  labs(title = "Age and Race: Who Is Most at Risk?",
       subtitle = "Distribution of police shooting victims by age group and race",
       x = "Age Group",
       y = "Number of Deaths",
       caption = "Source: Ian T. Adams | Data: Mapping Police Violence") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## Behavioral Context

### Armed Status

```{r armed-status, fig.cap="Armed/unarmed status of police shooting victims"}
armedinfo <- df_shootings %>%
  mutate(armed_status = replace_na(armedunarmed_status, "Unknown")) %>%
  group_by(armed_status) %>%
  summarise(n = n()) %>%
  arrange(desc(n)) %>%
  mutate(perc = scales::percent(n / sum(n), accuracy = .1))

ggplot(data = armedinfo, aes(x = n, y = reorder(armed_status, n))) +
  geom_col(aes(fill = n)) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_fill_gradient(low = "royalblue3", high = "navyblue") +
  labs(y = NULL, x = "Number of deaths",
       title = "Armed Status in Police Fatal Shootings",
       subtitle = paste(min_year, "-", latest_year),
       caption = "Source: Ian T. Adams | Data: Mapping Police Violence") +
  geom_text(aes(label = perc), hjust = -.25, size = 4, fontface = "bold") +
  coord_cartesian(clip = "off")
```

```{r unarmed-by-race, fig.cap="Percentage of victims who were unarmed, by race"}
unarmed_race <- df_shootings %>%
  filter(!is.na(armedunarmed_status) & !is.na(race_clean)) %>%
  mutate(unarmed = str_detect(tolower(armedunarmed_status), "unarmed")) %>%
  group_by(race_clean, unarmed) %>%
  summarize(count = n(), .groups = 'drop') %>%
  group_by(race_clean) %>%
  mutate(total = sum(count),
         pct = count / total * 100) %>%
  filter(unarmed == TRUE & race_clean %in% c("White", "Black", "Hispanic", "Asian"))

ggplot(unarmed_race, aes(x = reorder(race_clean, -pct), y = pct, fill = race_clean)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(pct, 1), "%\n(n=", count, ")")),
            vjust = -0.5, fontface = "bold", size = 4) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Who Gets Shot When Unarmed?",
       subtitle = "Percentage of police shooting victims who were unarmed, by race",
       x = NULL,
       y = "Percentage Unarmed",
       caption = "Source: Ian T. Adams | Data: Mapping Police Violence") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold")) +
  ylim(0, max(unarmed_race$pct) * 1.2)
```

### Mental Health Context

```{r mental-health-trend, fig.cap="Percentage of victims with documented mental illness symptoms over time"}
mental_trend <- df_shootings %>%
  filter(!is.na(symptoms_of_mental_illness)) %>%
  mutate(has_mental_illness = case_when(
    str_detect(tolower(symptoms_of_mental_illness), "yes|drug|alcohol") ~ "Yes",
    TRUE ~ "No/Unknown"
  )) %>%
  group_by(year, has_mental_illness) %>%
  summarize(count = n(), .groups = 'drop') %>%
  group_by(year) %>%
  mutate(total = sum(count),
         pct = count / total * 100) %>%
  filter(has_mental_illness == "Yes")

ggplot(mental_trend, aes(x = year, y = pct)) +
  geom_line(color = "darkblue", linewidth = 1.5) +
  geom_point(color = "darkblue", size = 3) +
  geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "dashed") +
  labs(title = "Mental Health Crisis: Growing Share of Police Shootings",
       subtitle = "Percentage of victims with documented mental illness symptoms",
       x = "Year",
       y = "Percentage with Mental Illness Symptoms",
       caption = "Source: Ian T. Adams | Data: Mapping Police Violence") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold")) +
  scale_x_continuous(breaks = seq(min(mental_trend$year), max(mental_trend$year), by = 2))
```

---

## Temporal Patterns

### When Do Shootings Occur?

```{r seasonal-heatmap, fig.width=10, fig.height=6, fig.cap="Heatmap showing daily patterns of police shootings across the calendar year"}
# Create month-day heatmap
heatmap_data <- df_shootings %>%
  mutate(month_num = as.numeric(month),
         day_num = day(date)) %>%
  filter(!is.na(month_num) & !is.na(day_num)) %>%
  group_by(month_num, day_num) %>%
  summarize(count = n(), .groups = 'drop')

ggplot(heatmap_data, aes(x = day_num, y = month_num, fill = count)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient(low = "lightyellow", high = "darkred", name = "Deaths") +
  scale_y_reverse(breaks = 1:12,
                  labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                             "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")) +
  scale_x_continuous(breaks = c(1, 8, 15, 22, 29)) +
  labs(title = "When Do Police Shootings Occur?",
       subtitle = paste("Daily patterns across the calendar year,", min_year, "-", latest_year),
       x = "Day of Month",
       y = NULL,
       caption = "Source: Ian T. Adams | Data: Mapping Police Violence") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        panel.grid = element_blank())
```

```{r day-of-week, fig.cap="Distribution of police shootings by day of week"}
dow_data <- df_shootings %>%
  group_by(day) %>%
  summarize(count = n()) %>%
  mutate(pct = count / sum(count) * 100,
         expected = 100/7)  # If random, each day = 14.3%

ggplot(dow_data, aes(x = day, y = pct)) +
  geom_col(aes(fill = pct > expected), show.legend = FALSE) +
  geom_hline(yintercept = 100/7, linetype = "dashed", color = "red", linewidth = 1) +
  geom_text(aes(label = paste0(round(pct, 1), "%")), vjust = -0.5, fontface = "bold") +
  scale_fill_manual(values = c("steelblue", "darkred")) +
  labs(title = "Are Weekends More Dangerous?",
       subtitle = "Distribution of police shootings by day of week",
       x = NULL,
       y = "Percentage of All Shootings",
       caption = "Source: Ian T. Adams | Data: Mapping Police Violence\nRed line = expected if random (14.3%)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold")) +
  ylim(0, max(dow_data$pct) * 1.1)
```

---

## Geographic Distribution

```{r geographic-map, fig.width=12, fig.height=8, fig.cap="Geographic distribution of police fatal shootings across the United States"}
# Filter for valid coordinates
df_map <- df_shootings %>%
  filter(!is.na(latitude) & !is.na(longitude))

world <- ne_countries(scale = "medium", returnclass = "sf")
sites <- st_as_sf(df_map, coords = c("longitude", "latitude"),
                  crs = 4326, agr = "constant")

ggplot(data = world) +
  geom_sf() +
  geom_sf(data = sites, size = .5, shape = 10, fill = "darkred", alpha = 0.6) +
  coord_sf(xlim = c(-125, -66), ylim = c(24, 49), expand = TRUE) +
  labs(title = "Geographic Distribution of Police Fatal Shootings",
       subtitle = paste(min_year, "-", latest_year),
       caption = "Source: Ian T. Adams | Data: Mapping Police Violence") +
  theme_minimal()
```

```{r top-states, fig.cap="Top 15 states by number of police fatal shootings"}
stateinfo <- df_shootings %>%
  filter(!is.na(state)) %>%
  group_by(state) %>%
  summarise(n = n()) %>%
  arrange(desc(n)) %>%
  top_n(15) %>%
  mutate(state = factor(state, levels = rev(unique(state))))

ggplot(stateinfo) +
  aes(x = state, weight = n) +
  geom_bar(fill = "steelblue") +
  labs(y = "Number of deaths",
       title = "Top 15 States for Police Fatal Shootings",
       subtitle = paste(min_year, "-", latest_year),
       caption = "Source: Ian T. Adams | Data: Mapping Police Violence") +
  coord_flip() +
  theme_minimal()
```

```{r top-cities, fig.width=10, fig.height=8, fig.cap="Top 20 cities by number of shootings, colored by percentage of victims who were unarmed"}
city_analysis <- df_shootings %>%
  filter(!is.na(city) & !is.na(state)) %>%
  mutate(city_state = paste0(city, ", ", state)) %>%
  group_by(city_state) %>%
  summarize(
    count = n(),
    unarmed_pct = sum(str_detect(tolower(armedunarmed_status), "unarmed"), na.rm = TRUE) / n() * 100,
    .groups = 'drop'
  ) %>%
  arrange(desc(count)) %>%
  head(20) %>%
  mutate(city_state = factor(city_state, levels = rev(city_state)))

ggplot(city_analysis, aes(x = count, y = city_state)) +
  geom_col(aes(fill = unarmed_pct)) +
  geom_text(aes(label = count), hjust = -0.2, size = 3) +
  scale_fill_gradient(low = "lightblue", high = "darkred",
                      name = "% Unarmed") +
  labs(title = "Top 20 Cities for Police Shootings",
       subtitle = paste("Color indicates percentage of victims who were unarmed,", min_year, "-", latest_year),
       x = "Number of Deaths",
       y = NULL,
       caption = "Source: Ian T. Adams | Data: Mapping Police Violence") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold")) +
  coord_cartesian(clip = "off")
```

---

## Per Capita Analysis

```{r per-capita-setup}
# Manual population data (with projections for 2024-2025)
pop_data <- tribble(
  ~year, ~estimate,
  2013, 316128839,
  2014, 318857056,
  2015, 320738994,
  2016, 323071755,
  2017, 325084756,
  2018, 326687501,
  2019, 328239523,
  2020, 331449281,
  2021, 331893745,
  2022, 333287557,
  2023, 334914895,
  2024, 336673595,  # Projection
  2025, 338289857   # Projection
)
```

```{r per-capita-rate, fig.cap="Police shooting deaths per capita by year, with 92% counting accuracy assumption"}
# Calculate per capita rates
events_normalized <- df_shootings %>%
  group_by(year) %>%
  summarize(total_events = n()) %>%
  left_join(pop_data, by = "year") %>%
  filter(!is.na(estimate)) %>%
  mutate(
    events_per_capita = total_events / estimate * 1000000,
    se = events_per_capita / sqrt(0.92 * total_events)
  )

ggplot(events_normalized, aes(x = year, y = events_per_capita)) +
  geom_col(color = "white", linewidth = 0.2, fill = "#7777AB") +
  geom_errorbar(aes(ymin = events_per_capita, ymax = events_per_capita + 1.96 * se),
                width = 0.2, color = "black", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.2f", events_per_capita), y = events_per_capita/2),
            vjust = 1.5) +
  labs(title = "US Police Shooting Deaths per 1,000,000 people by year",
       subtitle = "assuming 92% counting accuracy",
       y = "Fatal shootings per capita",
       x = "Year",
       caption = "Source: Ian T. Adams | Data: Mapping Police Violence\nPopulation: U.S. Census Bureau") +
  theme_bw() +
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14),
        plot.title = element_text(size = 14, face = "bold"),
        legend.position = "none") +
  scale_x_continuous(breaks = events_normalized$year, labels = events_normalized$year)
```

---

## Accountability and Socioeconomic Context

```{r accountability, fig.cap="Percentage of police shooting cases resulting in criminal charges", eval=("criminal_charges" %in% names(df_shootings))}
# Only run if criminal_charges column exists
if ("criminal_charges" %in% names(df_shootings)) {
  charges_trend <- df_shootings %>%
    filter(!is.na(criminal_charges)) %>%
    mutate(charged = !str_detect(tolower(criminal_charges), "no charges|no known|pending")) %>%
    group_by(year) %>%
    summarize(
      total = n(),
      charged = sum(charged),
      pct_charged = charged / total * 100,
      .groups = 'drop'
    )

  ggplot(charges_trend, aes(x = year, y = pct_charged)) +
    geom_line(color = "darkblue", linewidth = 1.5) +
    geom_point(aes(size = total), color = "darkblue", alpha = 0.6) +
    geom_hline(yintercept = mean(charges_trend$pct_charged),
               linetype = "dashed", color = "red") +
    scale_size_continuous(name = "Total Cases", range = c(2, 8)) +
    labs(title = "The Accountability Gap",
         subtitle = "Percentage of police shooting cases resulting in criminal charges",
         x = "Year",
         y = "Percentage Charged",
         caption = "Source: Ian T. Adams | Data: Mapping Police Violence\nRed line = average across all years") +
    theme_minimal() +
    theme(plot.title = element_text(size = 14, face = "bold")) +
    scale_x_continuous(breaks = seq(min(charges_trend$year), max(charges_trend$year), by = 2))
}
```

```{r income-inequality, fig.cap="Police shootings by neighborhood median income and race", eval=("median_household_income_acs_census_tract" %in% names(df_shootings))}
# Only run if income column exists
if ("median_household_income_acs_census_tract" %in% names(df_shootings)) {
  income_analysis <- df_shootings %>%
    filter(!is.na(median_household_income_acs_census_tract)) %>%
    mutate(
      income_bracket = cut(median_household_income_acs_census_tract,
                           breaks = c(0, 35000, 50000, 75000, 100000, 250001),
                           labels = c("<$35k", "$35k-50k", "$50k-75k", "$75k-100k", ">$100k"))
    ) %>%
    group_by(income_bracket, race_clean) %>%
    summarize(count = n(), .groups = 'drop') %>%
    filter(race_clean %in% c("White", "Black", "Hispanic"))

  ggplot(income_analysis, aes(x = income_bracket, y = count, fill = race_clean)) +
    geom_col(position = "dodge") +
    scale_fill_brewer(palette = "Set1", name = "Race") +
    labs(title = "Police Shootings and Neighborhood Poverty",
         subtitle = "Deaths by census tract median household income and race",
         x = "Neighborhood Median Income",
         y = "Number of Deaths",
         caption = "Source: Ian T. Adams | Data: Mapping Police Violence & U.S. Census Bureau") +
    theme_minimal() +
    theme(plot.title = element_text(size = 14, face = "bold"),
          axis.text.x = element_text(angle = 45, hjust = 1))
}
```

---

## Data Sources and Methods

### Primary Data Source

All data comes from [Mapping Police Violence](https://mappingpoliceviolence.us/), which aggregates information from:

- **Fatal Encounters** - A database documenting all deaths through police interaction
- **The Washington Post's Fatal Force database** - Tracking fatal police shootings
- **Additional cases** identified through crowdsourcing and independent research

### Update Frequency

This blog post **automatically downloads the latest MPV data whenever my website is rebuilt**. The database is continuously updated as new incidents are reported and verified.

### Important Methodological Considerations

1. **Incomplete Coverage**: This data captures police **killings** only, not all uses of lethal force. Research indicates that a slight majority of police shootings are survived, meaning these statistics undercount total lethal force incidents.

2. **Descriptive Statistics**: All analyses presented here are descriptive. Causal interpretation requires controlling for:
   - Local crime rates and patterns
   - Police deployment density
   - Population demographics
   - Socioeconomic factors
   - Jurisdictional policies and training

3. **Data Quality**: While MPV represents the most comprehensive publicly available database, it relies on media reports and public records, which may have gaps or inconsistencies.

4. **Racial Categories**: Racial/ethnic categorizations follow the source data and may not capture the full complexity of identity.

### Code Availability

The R code for this analysis is embedded in this R Markdown document. The source code is available in the [academic-website repository](https://github.com/ian-adams/academic-website).

---

## Related Resources

For an interactive version of this analysis with real-time filtering and exploration capabilities, see the [Interactive MPV Dashboard](/post/2026-01-04-mapping-police-violence-dashboard/).

---

*Last updated: `r formatted_date`*

*This analysis is provided for research and educational purposes. The data should be interpreted within the broader context of research on policing.*
