---
title: "Fun With the Scholar Package"
subtitle: ''
date: '2021-01-11T19:55:29-07:00'
authors: []
tags: []
categories: [R]
summary: 'Playing around with the updated `scholar` package.'
lastmod: '2021-01-11T19:55:29-07:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

I saw the `scholar` package [has a new maintainer on Github](https://github.com/jkeirstead/scholar), so thought I'd do a quick run through of what's available in the vignette. I was happy to see some updates, I think this is one of those fun but useful packages for people to learn about. Plus, it lets everyone make a little study of their favorite little solipsistic research subjects.

Make sure you have the updated version of `scholar`.

```{r scholar, eval=TRUE, include=TRUE}

# install.package("scholar")

library(scholar)

```

Setup the basic information by changing up the `id` argument, just visit [Google Scholar](https://scholar.google.com), click on the "My Profile" link, and copy the last character string from the url (minus everything after the & sign). Paste that to the `id` object below!

```{r id, echo=TRUE, eval=TRUE}
# Define the id for author
id <- 'g9lY5RUAAAAJ'

# Get profile and print name
l <- get_profile(id)
l$name 

# Get his citation history, i.e. citations to his work in a given year 
get_citation_history(id)

# Get his publications (a large data frame)
get_publications(id)

# Get number of articles
get_num_articles(id)

# Number of different journals published in
get_num_distinct_journals(id)

# Retrieve year of oldest publication
get_oldest_article(id)

# Number of publications in "top" journals
get_num_top_journals(id)

```

Need a little dose of imposter syndrome today? Dr. Hawking had 497 citations *in his first career year*. You can compare scholars based on their `id`. 

```{r compare, echo=TRUE}
# Compare yourself and Stephen Hawking
ids <- c('g9lY5RUAAAAJ', 'qj74uXkAAAAJ') 

# Get a data frame comparing the number of citations to their work in
# a given year 
compare_scholars(ids)

# Compare their career trajectories, based on year of first citation
compare_scholar_careers(ids)


```


Want to feel bad about yourself (part 2)? Use the [prediction algorithm from Acuna et al.](https://www.nature.com/nature/articles/489201a) to see where you'll be when you finally land that adjunct position!


```{r predict}
## Predict h-index of original method author, Daniel Acuna
id <- 'g9lY5RUAAAAJ'
predict <- predict_h_index(id)

plot(predict)
```

